---
title: "Greedy GP Sensor Placement — Algorithm Explanations"
subtitle: "Based on Krause, Singh & Guestrin (2008)  \n*Near-Optimal Sensor Placements
  in Gaussian Processes*  \nJMLR 9, 235–284\n"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: flatly
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
source("translation_python_testing.r")
```

---

# Background: Gaussian Processes and the Sensor Placement Problem

## Gaussian Processes

A **Gaussian Process** (GP) is a probability distribution over functions. A zero-mean
GP is fully characterised by its **kernel** (covariance function)
$k : \mathcal{X} \times \mathcal{X} \to \mathbb{R}$, written

$$f \sim \mathcal{GP}(0,\, k).$$

For a finite set of candidate locations $V = \{x_1, \ldots, x_n\} \subset \mathcal{X}$
the joint distribution is multivariate Gaussian:

$$f_V = (f(x_1), \ldots, f(x_n))^\top \sim \mathcal{N}(0,\, K)$$

where the **kernel matrix** $K \in \mathbb{R}^{n \times n}$ has entries $K_{ij} = k(x_i, x_j)$.

## Conditional Distribution

For any partition of $V$ into a selected set $S$ and its complement $\bar{S} = V \setminus S$,
the GP conditional is also Gaussian:

$$f_{\bar{S}} \mid f_S \sim \mathcal{N}\!\left(K_{\bar{S}S}\, K_{SS}^{-1} f_S,\;\;
  K_{\bar{S}\bar{S}} - K_{\bar{S}S}\, K_{SS}^{-1}\, K_{S\bar{S}}\right).$$

The **conditional variance** of a single point $x_k \notin S$ is therefore

$$\sigma^2(x_k \mid X_S)
  = k(x_k, x_k) - k_{kS}\, K_{SS}^{-1}\, k_{Sk}$$

where $k_{kS} = [k(x_k, x_i)]_{x_i \in S}$ is the cross-covariance row vector.

## The Sensor Placement Problem

Given a budget of $s$ sensors to place among $n$ candidate sites, we seek a set
$S^* \subset V,\, |S^*| = s$ that maximises some information-theoretic criterion.
The three criteria implemented here are:

| Criterion | Objective |
|---|---|
| Maximum entropy | $\max_S H(X_S)$ |
| Maximum mutual information | $\max_S I(X_S;\, X_{V \setminus S})$ |
| Minimum Nystrom KL divergence | $\min_S \mathrm{KL}(\mathrm{GP}_\text{full} \| \mathrm{GP}_\text{Nystrom}(S))$ |

All three are submodular functions of $S$, which means greedy selection provides a
$(1 - 1/e)$-approximation guarantee of the optimum.

---

# Helper / Utility Functions

## Matrix Inversion: `mat_inv`

For a symmetric positive definite (SPD) matrix $M$, the Cholesky factorisation
$M = U^\top U$ (with $U$ upper triangular) allows inversion as:

$$M^{-1} = U^{-1}\, (U^{-\top})$$

R's `chol2inv(chol(M))` implements this. The cost is $\approx n^3/3$ flops for
the factorisation plus $\approx n^3/3$ for the inversion, roughly half the cost of
a general LU-based inversion ($\approx n^3$ flops).

## Triangular Solve: `solve_tri`

Solves $Lx = b$ (or $L^\top x = b$) for lower or upper triangular $L$.
Triangular solves cost $O(n^2)$ and are the building block for all efficient
kernel computations in this file.

## Quadratic Form: `quad_form`

Computes the **quadratic form** $x_i^\top \Theta^{-1} x_i$ for each column $x_i$
of a matrix $X \in \mathbb{R}^{p \times q}$.

**Algorithm.** Write $\Theta = U^\top U$ (upper Cholesky). Then:

$$x^\top \Theta^{-1} x = x^\top U^{-1} U^{-\top} x = \| U^{-\top} x \|^2$$

Solve $U^\top t = x$ by forward substitution, then return $\|t\|^2$.
This is cheaper than explicitly forming $\Theta^{-1}$ first.

In the entropy algorithms this computes the **reduction in variance** from
conditioning on the selected set:

$$\sigma^2(x_k \mid X_S) = k(x_k, x_k) - \underbrace{k_{kS}\, K_{SS}^{-1}\, k_{Sk}}_{\text{quad\_form}(K_{SS},\, k_{kS}^\top)}$$

## Rank-1 Cholesky Downdate: `chol_downdate`

Given a lower Cholesky factor $L$ of $A = LL^\top$, and a vector $u$,
computes the lower Cholesky factor of the **rank-1 downdate**:

$$A - uu^\top = \tilde{L}\tilde{L}^\top$$

**Algorithm: Givens rotations.**  Process columns $i = 1, \ldots, n$ sequentially.
At column $i$, the current diagonal entry $L_{ii}$ and carry entry $u_i$ define a
2D rotation that zeros out $u_i$ while updating the new diagonal:

$$\tilde{L}_{ii} = \sqrt{L_{ii}^2 - u_i^2}$$
$$c_1 = \frac{L_{ii}}{\tilde{L}_{ii}}, \qquad c_2 = \frac{u_i}{\tilde{L}_{ii}}$$

Column $i$ and the carry vector $u$ are then updated:

$$L_{:,i} \leftarrow c_1 L_{:,i} - c_2\, u$$
$$u \leftarrow \frac{u}{c_1} - \frac{c_2}{c_1} L_{:,i}^{\text{new}}$$

After all $n$ columns the result $L$ is the Cholesky factor of $A - uu^\top$.

This operation is used in `mi_chol` to remove points from the precision Cholesky
as they are added to the selected set $S$.

---

# Greedy Maximum Entropy Selection

## Criterion

The differential entropy of a multivariate Gaussian $X_S \sim \mathcal{N}(0, K_{SS})$ is:

$$H(X_S) = \frac{1}{2} \log \det K_{SS} + \frac{|S|}{2}(1 + \log 2\pi)$$

Greedily maximising $H(X_S)$ is equivalent to greedily maximising $\log \det K_{SS}$.
By the chain rule of entropy, the greedy step reduces to:

$$x_k^* = \arg\max_{x_k \in V \setminus S}\; H(x_k \mid X_S)
        = \arg\max_{x_k \in V \setminus S}\; \sigma^2(x_k \mid X_S)$$

i.e. always add the point with the **largest conditional variance**.

## Algorithm 1 (Greedy Max-Entropy)

$$\boxed{
\begin{array}{l}
\textbf{Input: } \text{candidate set } V,\, \text{budget } s \\
S \leftarrow \emptyset \\
\textbf{for } i = 1, \ldots, s: \\
\quad k^* \leftarrow \arg\max_{k \in V \setminus S}\; \sigma^2(x_k \mid X_S) \\
\quad S \leftarrow S \cup \{k^*\} \\
\textbf{return } S
\end{array}
}$$

The conditional variance update at each step uses:

$$\sigma^2(x_k \mid X_S) = K_{kk} - K_{kS}\, K_{SS}^{-1}\, K_{Sk}$$

The four implementations below differ only in **how they maintain** $K_{SS}^{-1}$
(or an equivalent factorisation) across iterations.

## `entropy_naive` — $O(ns^3 + s^4)$

At each of the $s$ steps, the submatrix $K_{SS}$ is passed to `quad_form`, which
calls `chol()` from scratch. This costs $O(|S|^3)$ per iteration, giving total
cost $O(ns^3 + s^4)$.

## `entropy_prec` — $O(ns^2)$

Maintains the **precision matrix** $P = K_{SS}^{-1}$ and updates it by the
**Schur complement block inversion formula** after adding each point.

When point $k$ is added to $S$, the new joint covariance is:
$$K_{S \cup \{k\},\, S \cup \{k\}} = \begin{pmatrix} K_{SS} & k_{Sk} \\ k_{kS} & K_{kk} \end{pmatrix}$$

By the block matrix inversion lemma, the new precision is:

$$P_{\text{new}} = \begin{pmatrix}
  P + w\, v v^\top & -w\, v \\
  -w\, v^\top      & w
\end{pmatrix}$$

where

$$w = \frac{1}{\sigma^2(x_k \mid X_S)}, \qquad v = P\, K_{Sk}.$$

The conditional variance update for all remaining points reduces to a
**rank-1 subtraction**:

$$\sigma^2(x_j \mid X_{S \cup \{k\}}) = \sigma^2(x_j \mid X_S)
  - \frac{\mathrm{Cov}^2(x_j, x_k \mid X_S)}{\sigma^2(x_k \mid X_S)}$$

where $\mathrm{Cov}(x_j, x_k \mid X_S) = K_{jk} - K_{jS}\, K_{SS}^{-1} K_{Sk}$.

## `entropy_prechol` — $O(ns^2)$

Instead of the precision matrix, maintains the lower Cholesky factor $L$ of
$K_{SS}$, where $K_{SS} = LL^\top$. Uses the **"up-looking" Cholesky update**:
when the new point $k$ is added, the new Cholesky factor satisfies

$$\begin{pmatrix} K_{SS} & k_{Sk} \\ k_{kS} & K_{kk} \end{pmatrix}
= \begin{pmatrix} L & 0 \\ \ell^\top & d \end{pmatrix}
  \begin{pmatrix} L^\top & \ell \\ 0 & d \end{pmatrix}$$

Solving $L\ell = K_{Sk}$ (size $|S|$ forward substitution) gives $\ell$, and
the new diagonal entry is $d = \sqrt{K_{kk} - \|\ell\|^2}$.

## `entropy_chol` — $O(ns^2)$

The most cache-friendly variant. Maintains an $n \times s$ matrix $L$ where
column $i$ is the $i$-th column of the lower Cholesky factor of the **full**
kernel matrix $K$, restricted to the selected set.

**Left-looking Cholesky column update** at step $i$:

$$L_{:,i} \leftarrow \frac{K_{:,k} - L_{:,1:i-1}\, L_{k,1:i-1}^\top}{\sqrt{L_{k,i}}}$$

The conditional variance update is simply:

$$\sigma^2(x_j \mid X_{S \cup \{k\}}) = \sigma^2(x_j \mid X_S) - L_{j,i}^2$$

**Demonstration** that all four variants agree:

```{r entropy-demo}
# RBF (squared-exponential) kernel with lengthscale = 1
rbf <- function(X1, X2) {
  n1 <- nrow(X1); n2 <- nrow(X2)
  d2 <- as.matrix(dist(rbind(X1, X2)))[1:n1, (n1+1):(n1+n2), drop=FALSE]
  exp(-d2^2 / 2)
}
rbf_diag <- function(X) rep(1, nrow(X))

set.seed(42)
X <- matrix(rnorm(30), ncol = 2)  # 15 points in 2D

data.frame(
  variant = c("entropy_naive", "entropy_prec", "entropy_prechol", "entropy_chol"),
  selected_indices = c(
    paste(entropy_naive(X, rbf, rbf_diag, 4),    collapse = " "),
    paste(entropy_prec(X, rbf, rbf_diag, 4),     collapse = " "),
    paste(entropy_prechol(X, rbf, rbf_diag, 4),  collapse = " "),
    paste(entropy_chol(X, rbf, rbf_diag, 4),     collapse = " ")
  )
)
```

---

# Greedy Maximum Mutual Information Selection

## Criterion

Mutual information between the selected set $X_S$ and the unselected set
$X_{\bar{S}} = X_{V \setminus S}$ is:

$$I(X_S;\, X_{\bar{S}}) = H(X_S) + H(X_{\bar{S}}) - H(X_V)
  = \frac{1}{2}\log \frac{\det K_{SS} \cdot \det K_{\bar{S}\bar{S}}}{\det K}$$

For Gaussian variables, MI decomposes using the chain rule. Adding point $k$ to $S$
gives the greedy score:

$$\Delta_k = I(x_k;\, X_{V \setminus (S \cup \{k\})} \mid X_S)
           = H(x_k \mid X_S) - H(x_k \mid X_{V \setminus S})$$

For Gaussians entropy is a log:

$$\Delta_k = \frac{1}{2}\log \sigma^2(x_k \mid X_S) - \frac{1}{2}\log \sigma^2(x_k \mid X_{V \setminus S})$$

Maximising over $k$ is equivalent to maximising the **product**:

$$x_k^* = \arg\max_{k \in V \setminus S}\;
  \underbrace{\sigma^2(x_k \mid X_S)}_{\text{cond\_var1}[k]}
  \cdot
  \underbrace{\sigma^2(x_k \mid X_{V \setminus S})}_{\text{cond\_var2}[k]}$$

**Why is cond_var2 the diagonal of the precision?**  
The **full conditional** variance of $x_k$ given all other points in $\bar{S}$ is
related to the precision matrix $P_{\bar{S}} = K_{\bar{S}\bar{S}}^{-1}$ by:

$$\sigma^2(x_k \mid X_{\bar{S} \setminus \{k\}}) = \frac{1}{[K_{\bar{S}\bar{S}}^{-1}]_{kk}}$$

This is the **Schur complement** property: the $k$-th diagonal of the precision
equals the reciprocal of the full conditional variance of the $k$-th variable.

## Algorithm 2 (Greedy Max-MI)

$$\boxed{
\begin{array}{l}
\textbf{Input: } V,\, s \\
S \leftarrow \emptyset,\quad P \leftarrow K^{-1} \\
\textbf{for } i = 1, \ldots, s: \\
\quad k^* \leftarrow \arg\max_{k \in V \setminus S}\;
  \sigma^2(x_k \mid X_S) \cdot [P_{V \setminus S}]_{kk}^{-1} \\
\quad S \leftarrow S \cup \{k^*\} \\
\quad P_{V \setminus S} \leftarrow \text{remove } k^* \text{ from } P \text{ (Schur complement)} \\
\textbf{return } S
\end{array}
}$$

## Schur Complement Update for the Precision (used in `mi_prec`)

When point $k$ (at position $j$ in the current candidate vector) is moved from
$\bar{S}$ to $S$, the precision matrix of the remaining candidates updates as:

$$P_{\text{new}} = \left(P_{\bar{S} \setminus \{k\}, \bar{S} \setminus \{k\}}\right)^{-1 \leftarrow \text{this IS the update}}
= P_{-j,-j} - \frac{P_{-j,j}\, P_{j,-j}}{P_{j,j}}$$

where $-j$ denotes all rows/columns except $j$.
This corresponds to **marginalising** $x_k$ out of the precision: conditioning in
covariance is marginalising in precision, and vice versa.

## `mi_naive` — $O(n^4 s)$

Recomputes both conditional variances from scratch every iteration. Each outer step
costs $O(n^3)$ for inverting $K_{\bar{S}\bar{S}}$.

## `mi_prec` — $O(n^3)$

Initialises $P = K^{-1}$ once ($O(n^3)$), then maintains $P_{\bar{S}}$ using the
Schur complement deletion above ($O(n^2)$ per step). Also maintains a left-looking
Cholesky for `cond_var1` (same as `entropy_chol`).

## `mi_chol` — $O(n^3)$

Avoids storing the full $n \times n$ precision matrix by maintaining its lower
Cholesky factor $L_2$ where $K^{-1} = L_2^\top L_2$.

**Initialisation of $L_2$.**  
Let $J$ be the flip permutation matrix ($J_{ij} = 1$ if $i + j = n+1$, 0 otherwise).
Then $K_\text{rev} = JKJ$ (the kernel matrix with rows and columns reversed).
Compute $K_\text{rev} = U^\top U$ via Cholesky. The inverse of $K$ can be
obtained without explicit inversion:

$$K^{-1} = J (U^\top U)^{-1} J = J U^{-1} U^{-\top} J
= (J U^{-\top} J)^\top (J U^{-\top} J)$$

Setting $L_2 = J U^{-\top} J$ (flip $U^{-\top}$ in both dimensions) gives a valid
lower Cholesky of $K^{-1}$:

$$L_2^\top L_2 = K^{-1}$$

**Diagonal entries** give `cond_var2`:

$$[K^{-1}]_{kk} = \| L_2[k,\:] \|^2 = \texttt{rowSums}(L_2 * L_2)[k]$$

**Downdate of $L_2$.**  
When candidate $k$ (at position $j$) is selected, the precision of the remaining
candidates must be updated:

$$K_{\bar{S} \setminus \{k\}, \bar{S} \setminus \{k\}}^{-1}
= K_{\bar{S}\bar{S}}^{-1} - \frac{K_{\bar{S}\bar{S}}^{-1} e_j\, e_j^\top K_{\bar{S}\bar{S}}^{-1}}
  {[K_{\bar{S}\bar{S}}^{-1}]_{jj}}$$

Defining $u = K_{\bar{S}\bar{S}}^{-1} e_j \,/\, \sqrt{[K_{\bar{S}\bar{S}}^{-1}]_{jj}}$
this is the rank-1 downdate $L_2 L_2^\top - u u^\top$, handled by `chol_downdate`.

**Demonstration** that all three MI variants agree:

```{r mi-demo}
data.frame(
  variant = c("mi_naive", "mi_prec", "mi_chol"),
  selected_indices = c(
    paste(mi_naive(X, rbf, rbf_diag, 4), collapse = " "),
    paste(mi_prec(X, rbf, rbf_diag, 4),  collapse = " "),
    paste(mi_chol(X, rbf, rbf_diag, 4),  collapse = " ")
  )
)
```

---

# Greedy Nystrom Ordering

## Criterion

The **Nystrom approximation** to a GP using inducing set $S$ replaces the full
posterior covariance with a low-rank approximation. The quality loss is measured by
the **KL divergence**:

$$\mathrm{KL}\!\left(\mathrm{GP}_\text{full} \;\|\; \mathrm{GP}_\text{Nystrom}(S)\right)
= \frac{1}{2} \sum_{j \notin S} \log \frac{\sigma^2(x_j)}{\sigma^2(x_j \mid X_S)}$$

where $\sigma^2(x_j)$ is the marginal (prior) variance and $\sigma^2(x_j \mid X_S)$
is the posterior variance after conditioning on $S$.

This criterion is non-negative (since $\sigma^2(x_j \mid X_S) \le \sigma^2(x_j)$)
and equals zero when $S = V$.

## Greedy algorithm

At each step, add the point $k^*$ that **most reduces** the KL:

$$k^* = \arg\min_{k \in V \setminus S}\;
  \Delta\mathrm{KL}(S, k) = \arg\min_{k}\;
  \sum_{\substack{j \notin S \\ j \neq k}}
  \log \frac{\sigma^2(x_j \mid X_S)}{\sigma^2(x_j \mid X_{S \cup \{k\}})}$$

The inner sum uses the fact that after adding $k$, the conditional variance of every
remaining point updates as:

$$\sigma^2(x_j \mid X_{S \cup \{k\}})
= \sigma^2(x_j \mid X_S)
  - \frac{\mathrm{Cov}^2(x_j, x_k \mid X_S)}{\sigma^2(x_k \mid X_S)}$$

## `nystrom_order_naive` — $O(n^2 s^2)$

For each candidate $k$ at each outer step, tentatively computes the full conditional
variance update vector (cost $O(n)$ per candidate) and accumulates the log-ratio sum.
This inner loop over $n$ candidates costs $O(n^2)$ per outer step.

## `nystrom_order` — $O(n^2 s)$

Maintains the full **conditional covariance matrix**
$C = K - K_{:,S}\, K_{SS}^{-1}\, K_{S,:}$ (size reduces by one each step).
All candidate scores can be computed simultaneously from a single matrix operation.

For candidate $k$ (column/row $j$ in the current $C$), the score matrix is:

$$\text{scores}[a, k] = \underbrace{C_{aa}}_{\sigma^2(x_a \mid X_S)}
  - \frac{C_{ak}^2}{C_{kk}}$$

The full KL score for candidate $k$ is:

$$\Delta\mathrm{KL}(k) = \sum_{a \neq k} \log \frac{C_{aa}}{\text{scores}[a,k]}$$

The diagonal is set to $C_{aa}$ (a point does not affect its own variance), so the
score matrix can be safely log-transformed and summed column-wise.

After selecting $k$ (at position $j$), the conditional covariance shrinks via
a rank-1 downdate:

$$C_\text{new} = C_{-j,-j} - u\, u^\top, \quad u = \frac{C_{:,j}}{\sqrt{C_{jj}}}$$

**Demonstration** that both Nystrom variants agree:

```{r nystrom-demo}
data.frame(
  variant = c("nystrom_order_naive", "nystrom_order"),
  selected_indices = c(
    paste(nystrom_order_naive(X, rbf, rbf_diag, 4), collapse = " "),
    paste(nystrom_order(X, rbf, rbf_diag, 4),       collapse = " ")
  )
)
```

---

# Summary Table

| Function | Criterion | Strategy | Complexity | Key data structure |
|---|---|---|---|---|
| `entropy_naive` | Max entropy | Recompute $K_{SS}^{-1}$ each step | $O(ns^3 + s^4)$ | $s \times s$ covariance submatrix |
| `entropy_prec` | Max entropy | Maintain precision $K_{SS}^{-1}$ | $O(ns^2)$ | $s \times s$ precision matrix |
| `entropy_prechol` | Max entropy | Maintain Cholesky of $K_{SS}$ | $O(ns^2)$ | $s \times s$ Cholesky (up-looking) |
| `entropy_chol` | Max entropy | Left-looking Cholesky on all $n$ points | $O(ns^2)$ | $n \times s$ Cholesky columns |
| `mi_naive` | Max MI | Recompute both precisions each step | $O(n^4 s)$ | Full $n \times n$ kernel matrix |
| `mi_prec` | Max MI | Maintain precision of unselected | $O(n^3)$ | $n \times n$ shrinking precision |
| `mi_chol` | Max MI | Maintain Cholesky of precision | $O(n^3)$ | $n \times n$ shrinking Cholesky |
| `nystrom_order_naive` | Min Nystrom KL | Evaluate each candidate in a loop | $O(n^2 s^2)$ | $n \times s$ Cholesky columns |
| `nystrom_order` | Min Nystrom KL | Vectorised score matrix + rank-1 update | $O(n^2 s)$ | $n \times n$ conditional covariance |

**For practical use:**

- When $n$ is small and $s$ is comparable to $n$, any variant works; prefer the
  naive variants for simplicity.
- For large $n$ with small $s$, `entropy_chol` and `mi_chol` are the preferred choices.
- The Nystrom criterion explicitly optimises approximation quality rather than
  information, which can matter when the GP is used inside a downstream estimator.

---

# Kernel Interface

All nine functions share the same two-argument kernel interface:

```r
# kernel(X1, X2) : nrow(X1) × nrow(X2) covariance matrix
# kernel_diag(X) : length-nrow(X) marginal variance vector = diag(kernel(X, X))
```

`kernel_diag` exists for efficiency: for stationary kernels (RBF, Matérn) the
diagonal is constant, so $n$ evaluations suffice instead of the $n^2$ required by
`diag(kernel(X, X))`. This is critical for maintaining the $O(ns^2)$ complexity of
the efficient entropy variants.
